Spanner: Google's Globally-Distributed Database \newline Seminar Report

### 2016-02-25 paper/paper.tex
Remove tbr
Iterate section 1 and 2

Google already has two other highly scalable distributed databases, i.e.
Bigtable~\cite{Chang:2008} and Megastore~\cite{Baker:2011}. Hence, there must be
a reason, why they did not just use one of these but implemented another
distributed database. According to Google, it is difficult for applications/users
to deal with complex and evolving schemas in Bigtable. Moreover, Bigtable only
guarantees to be eventually-consistent. In contrast, Spanner provides much
stronger consistency guarantees. On the other hand, Megastore suffers from poor
write throughput even though many applications use it.

@@

\subsection{Spanserver Software Stack \tbr}
\label{subsec:spanserver-software-stack}

@@

On top of each tablet, a Paxos state machine is implemented in order to support
consistent replication. One Paxos state machine is elect as leader. Leader leases
are assigned in a long-lived manner (default lease length is 10 seconds). A leader
implements three additional layers. A \emph{lock table} which stores the state
for two-phase locking and therefore implements the concurrency control. On top
of the lock table, a \emph{transaction manager} is implemented to coordinate
distributed transactions. In turn, this transaction manager is used to implement
a \emph{participant leader}. This leader is only utilized if a transaction
involves multiple Paxos groups (set of replicas). In this case, one of the
participant groups is chosen as coordinator and the leader of this group is
referred to as \emph{coordinator leader}.
Figure~\ref{subfig:spanserver-software-stack} shows the spanserver software stack.

@@

All the data within a directory has the same replication configuration. As will
be described later, the concept of directories enables Google to improve the
performance of Spanner by placing directories which are frequently accessed
together in the same Paxos group. The fact that in general multiple directories
are placed in the same Paxos group is another difference between the key-value
mapping of Spanner and the one used in Bigtable. A Spanner tablet may contain
multiple contiguous ranges of keys (= directories) as opposed to a Bigtable tablet
which may only contain one contiguous range of keys.
Figure~\ref{fig:comparison-tablets} shows an example comparison of the two tablet
types.

@@

This partitioning of the row space is critical for the performance of Spanner
because it allows Google to physically \emph{co-locate} data that is frequently
accessed together. Therefore, directories are the smallest unit for which an
application/user can specify geographic-replication properties. An application
can control replication in two dimensions, namely the number and types of
replicas, and the geographic placement of replicas.

@@

There exists another layer on top of the directory-bucketed key-value mappings.
This layer is probably the most intuitive one for people who are used to common
database systems. The data model allows applications to create databases in a
universe each of which may contain any number of schematized
\emph{semi-relational} tables. A table has rows, columns and, as a third
dimension, versioned values. Spanner's tables are semi-relational because rows
must have names, i.e. every table must have an ordered set of one or more
primary-key columns. This is useful for applications because the locality of
data can be controlled by choosing keys accurately.

Therefore, Spanner requires the application/user to partition databases into
one or more hierarchies of tables. Based on this hierarchical information, Spanner
builds the directories. A directory contains hierarchically ordered rows from
all tables of a hierarchy which belong to the same key of the topmost table of
the hierarchy. Figure~\ref{subfig:interleaved-tables} shows an example of this
concept. In this example, the \texttt{Users} table is the topmost table of the
hierarchy and the \texttt{Albums} table is one level below. As we can see, all
rows of \texttt{Albums} connected to user \texttt{1} are located in the same
directory (\texttt{3365}). This directory has a common prefix, i.e. \texttt{1},
from the \texttt{Users} table. The same principle can be seen for key \texttt{2}
of the \texttt{Users} table. This interleaving of tables exploits locality
relationships of multiple tables and therefore is significant for Spanner's
performance.

@@

This section describes the key enabler of Spanner, i.e. the novel \emph{TrueTime}
API. Essentially, this API enables Google to assign timestamps to distributed
transactions which reflect serialization order.

TrueTime represents time as a $TTinterval$. A $TTinterval$ is an interval with a
bounded time uncertainty and has two endpoints, i.e. $earliest$ and $latest$,
which are of type $TTstamp$ (which is a timestamp in the context of Spanner).
Table~\ref{tbl:truetime-api} summarizes the three methods provided by the TrueTime
API. However, the most important method is $TT.now()$ which returns a $TTinterval$
which is guaranteed to contain the absolute time of the invocation of $TT.now()$,
i.e. $TT.now().earliest \leq t_{abs}(e) \leq TT.now().latest$, where $e$ denotes
the event of invoking $TT.now()$. More precisely, the absolute time of invocation
of $TT.now()$ is exactly in the middle this $TTinterval$. The interval itself
spans over $2\cdot\epsilon$, where $\epsilon$ denotes the instantaneous error
bound. $\epsilon$ is derived from applied worst-case local clock drift (assumed
to be $200\mu$s/sec). This instantaneous error bound is adjusted over time for
every single machine.

@@

The TrueTime API allows Spanner to guarantee some correctness properties of the
implemented concurrency control. These enable some of the main features, i.e.
externally-consistent transactions, lock-free read-only transactions and so-called
\emph{snapshot reads}. Snapshot reads are non-blocking reads in the past at a
specific timestamp $t$. In this case, Spanner guarantees that snapshot reads will
see the effects of all transactions committed as of $t$.

@@

Read-write transactions use two-phase locking and Spanner may only assign
timestamps in between the two phases. Spanner enforces another invariant, i.e. a
\emph{monotonicity invariant}, which makes use of the disjointness invariant.
This monotonicity invariant guarantees that within each Paxos group, Spanner
assigns timestamps to Paxos writes in monotonically increasing order. To guarantee
this across leaders, Spanner utilizes the disjointness invariant. This implies
that a leader must only assign timestamps within the interval of its leader
lease. $s_{max}$ is advanced to $s$ whenever a timestamp $s$ is assigned by a
leader.

@@

Figure~\ref{fig:read-write-transaction} illustrates the timeline of a distributed
read-write transaction. There are three participants groups involved, i.e. $P_2$,
$P_3$ and $P_{CL}$ (which is the coordinator leader). In the first step, all three
participants acquire their locks and execute $TT.now()$ right afterwards to obtain
timestamps independently. Every participant logs his timestamp through Paxos and
after this was done, sends its timestamp to the coordinator leader. The
coordinator leader then computes an overall timestamp for the transaction. This
timestamp must be greater than any timestamp the leader has assigned before,
greater or equal to all timestamps received from the participants and greater than
$TT.now().latest$ at the time the coordinator leader received the commit message.
Then, the computed timestamp is logged by the coordinator leader through Paxos
but no participants are notified until the commit wait rule is satisfied. After
commit wait is satisfied, the coordinator leader notifies all other participants
which are allowed to release their locks after receiving this notification.

@@

The safe time is determined by the safe time of a Paxos state machine, denoted
as $t_{safe}^{Paxos}$, and the safe time of the transaction manager, i.e.
$t_{safe}^{TM}$. The overall safe time of a replica then is the minimum of
these two.

@@

This section summarizes the core results of the evaluation of Spanner, i.e. the
latency and throughput performance as well as some results concerning availability
and the TrueTime API.

@@

A component that may influence the performance of Spanner is the distributed
file system, i.e. Colossus, the successor of the Google File System. The internet 
does not provide reliable information about the implementation or the performance
of Colossus. It is unknown how much impact this distributed file system has on the
performance of Spanner. Thus, it is hard to say if a reimplementation of Spanner
would be able to compete with Spanner's performance if a different distributed
file system is used.

Of course, a distributed database system may also be built upon other concepts,
e.g. the Network Time Protocol (NTP). But this would result in poor performance
compared to Spanner because the error interval would be much larger then the one
used by Spanner, i.e. $2\cdot\epsilon$. And besides providing distributed
general-purpose transactions, it is the performance of Spanner that impressives
the most. A system using NTP would never be able to achieve this performance
while guaranteeing serialization order of transactions. Spanner can synchronize
multiple systems without any communication between these systems. NTP-based
systems would have to communicate with each other. For systems far apart such a
NTP-based solution would never be able to compete with Spanner.

@@

It would also be interesting to know how failures of the two singletons per
universe, i.e. universemaster and placement driver, would affect Spanner. Since
the universemaster is just a debugging console, most probably a failure in this
component will not affect Spanner at all. This would probably be a totally
different story if the placement driver does not operate correctly. First of all,
there would be no load balancing between zones. Hence, zones could run out of
space if the placement driver is down long enough. Furthermore, updated
replication constraints, would not be propagated since the placement driver is
responsible to move the data if the replication constraints are modified.
Therefore I do not understand why there is no redundancy in placement driver.

>>>>

Google already has two other highly scalable distributed databases, i.e.,
Bigtable~\cite{Chang:2008} and Megastore~\cite{Baker:2011}. Hence, there must be
a reason, why they did not just use one of these but implemented another
distributed database. According to Google, it is difficult for applications/users
to deal with complex and evolving schemas in Bigtable. Moreover, Bigtable only
guarantees to be eventually-consistent. In contrast, Spanner provides much
stronger consistency guarantees. On the other hand, Megastore suffers from poor
write throughput even though many applications use it.

@@

\subsection{Spanserver Software Stack}
\label{subsec:spanserver-software-stack}

@@

On top of each tablet, a Paxos state machine is implemented in order to support
consistent replication. One Paxos state machine is elected as leader. Leader
leases are assigned in a long-lived manner (default lease length is 10 seconds).
A leader implements three additional layers. A \emph{lock table} which stores the
state for two-phase locking and therefore implements the concurrency control. On
top of the lock table, a \emph{transaction manager} is implemented to coordinate
distributed transactions. In turn, this transaction manager is used to implement
a \emph{participant leader}. This leader is only utilized if a transaction
involves multiple Paxos groups (set of replicas). In this case, one of the
participant groups is chosen as a coordinator and the leader of this group is
referred to as \emph{coordinator leader}.
Figure~\ref{subfig:spanserver-software-stack} shows the spanserver software stack.

@@

All the data within a directory has the same replication configuration. As will
be described later, the concept of directories enables Google to improve the
performance of Spanner by placing directories which are frequently accessed
together in the same Paxos group. The fact that in general multiple directories
are placed in the same Paxos group is another difference between the key-value
mapping of Spanner and the one used in Bigtable. A Spanner tablet may contain
multiple contiguous ranges of keys (= directories) with gaps between contiguous
ranges as opposed to a Bigtable tablet which may only contain one contiguous key
range. Figure~\ref{fig:comparison-tablets} shows an example comparison of the two
tablet types.

@@

This partitioning of the row space is critical for the performance of Spanner
because it allows Google to physically \emph{co-locate} data that is frequently
accessed together. Therefore, directories are the smallest unit for which an
application/user can specify geographic-replication properties. An application
can control replication along two dimensions, namely the number and types of
replicas, and the geographic placement of replicas.

@@

There exists another layer on top of the directory-bucketed key-value mappings,
i.e., schematized \emph{semi-relational} tables. This layer is probably the most
intuitive one for people who are used to common database systems. The data model
allows applications to create databases in a universe each of which may contain
an arbitrary number of schematized semi-relational tables. A table has rows,
columns, and (as a third dimension) versioned values. Spanner's tables are
semi-relational because rows must have names, i.e., every table must have an
ordered set of one or more primary-key columns. This is useful for applications
because the locality of data can be controlled by choosing keys accurately.

Therefore, Spanner requires the application/user to partition databases into
one or more hierarchies of tables. Based on this hierarchical information, Spanner
builds the directories. A directory contains hierarchically ordered rows from
all tables of a hierarchy which belong to the same key of the topmost table of
the hierarchy. Figure~\ref{subfig:interleaved-tables} shows an example of this
concept. In this example, the \texttt{Users} table is the topmost table of the
hierarchy and the \texttt{Albums} table is one level below. As we can see, all
rows of \texttt{Albums} connected to user \texttt{1} are located in the same
directory (\texttt{3365}). This directory has a common prefix, \texttt{1},
from the \texttt{Users} table. The same principle applies to key \texttt{2}
of the \texttt{Users} table. This interleaving of tables exploits locality
relationships of multiple tables and therefore is significant for Spanner's
performance.

@@

This section describes the key enabler of Spanner, i.e., the novel \emph{TrueTime}
API. Essentially, this API enables Google to assign timestamps to distributed
transactions which reflect serialization order.

TrueTime represents time as a $TTinterval$. A $TTinterval$ is an interval with a
bounded time uncertainty and has two endpoints, i.e., $earliest$ and $latest$,
which are of type $TTstamp$ (which is a timestamp in the context of Spanner).
Table~\ref{tbl:truetime-api} summarizes the three methods provided by the TrueTime
API. However, the most important method is $TT.now()$ which returns a $TTinterval$
which is guaranteed to contain the absolute time of the invocation of $TT.now()$,
i.e., $TT.now().earliest \leq t_{abs}(e) \leq TT.now().latest$, where $e$ denotes
the event of invoking $TT.now()$. More precisely, the absolute time of invocation
of $TT.now()$ is exactly in the middle this $TTinterval$. The interval itself
spans over $2\cdot\epsilon$, where $\epsilon$ denotes the instantaneous error
bound. $\epsilon$ is derived from applied worst-case local clock drift (assumed
to be $200\mu$s/sec). This instantaneous error bound is adjusted over time for
every single machine.

@@

The TrueTime API allows Spanner to guarantee some correctness properties of the
implemented concurrency control. These enable some of the main features, i.e.,
externally-consistent transactions, lock-free read-only transactions and so-called
\emph{snapshot reads}. Snapshot reads are non-blocking reads in the past at a
specific timestamp $t$. In this case, Spanner guarantees that snapshot reads will
see the effects of all transactions committed as of $t$.

@@

Read-write transactions use two-phase locking and Spanner may only assign
timestamps in between the two phases. Spanner enforces another invariant, i.e.,
a \emph{monotonicity invariant}, which makes use of the disjointness invariant.
This monotonicity invariant guarantees that within each Paxos group, Spanner
assigns timestamps to Paxos writes in monotonically increasing order. To guarantee
this across leaders, Spanner utilizes the disjointness invariant. This implies
that a leader must only assign timestamps within the interval of its leader
lease. $s_{max}$ is advanced to $s$ whenever a timestamp $s$ is assigned by a
leader.

@@

Figure~\ref{fig:read-write-transaction} illustrates the timeline of a distributed
read-write transaction. There are three participants groups involved, i.e., $P_2$,
$P_3$ and $P_{CL}$ (which is the coordinator leader). In the first step, all three
participants acquire their locks and execute $TT.now()$ right afterwards to obtain
timestamps independently. Every participant logs his timestamp through Paxos and
after this was done, sends its timestamp to the coordinator leader. The
coordinator leader then computes an overall timestamp for the transaction. This
timestamp must be greater than any timestamp the leader has assigned before,
greater or equal to all timestamps received from the participants and greater than
$TT.now().latest$ at the time the coordinator leader received the commit message.
Then, the computed timestamp is logged by the coordinator leader through Paxos
but no participants are notified until the commit wait rule is satisfied. After
commit wait is satisfied, the coordinator leader notifies all other participants
which are allowed to release their locks after receiving this notification.

@@

The safe time is determined by the safe time of a Paxos state machine, denoted
as $t_{safe}^{Paxos}$, and the safe time of the transaction manager, i.e.,
$t_{safe}^{TM}$. The overall safe time of a replica then is the minimum of
these two.

@@

This section summarizes the core results of the evaluation of Spanner, i.e., the
latency and throughput performance as well as some results concerning availability
and the TrueTime API.

@@

A component that may influence the performance of Spanner is the distributed
file system, i.e., Colossus, the successor of the Google File System. The internet 
does not provide reliable information about the implementation or the performance
of Colossus. It is unknown how much impact this distributed file system has on the
performance of Spanner. Thus, it is hard to say if a reimplementation of Spanner
would be able to compete with Spanner's performance if a different distributed
file system is used.

Of course, a distributed database system may also be built upon other concepts,
e.g. the Network Time Protocol (NTP). But this would result in poor performance
compared to Spanner because the error interval would be much larger then the one
used by Spanner, i.e., $2\cdot\epsilon$. And besides providing distributed
general-purpose transactions, it is the performance of Spanner that impressives
the most. A system using NTP would never be able to achieve this performance
while guaranteeing serialization order of transactions. Spanner can synchronize
multiple systems without any communication between these systems. NTP-based
systems would have to communicate with each other. For systems far apart such a
NTP-based solution would never be able to compete with Spanner.

@@

It would also be interesting to know how failures of the two singletons per
universe, i.e., universemaster and placement driver, would affect Spanner. Since
the universemaster is just a debugging console, most probably a failure in this
component will not affect Spanner at all. This would probably be a totally
different story if the placement driver does not operate correctly. First of all,
there would be no load balancing between zones. Hence, zones could run out of
space if the placement driver is down long enough. Furthermore, updated
replication constraints, would not be propagated since the placement driver is
responsible to move the data if the replication constraints are modified.
Therefore I do not understand why there is no redundancy in placement driver.
